---
title: "Making Pseudonyms in R"
output:
  github_document: default
  html_notebook: default
---



In this document we will show how to make pseudonyms in R. The methods presented here are discussed in the ENISA report "[Pseudonymisation techniques and best practices](https://www.enisa.europa.eu/publications/pseudonymisation-techniques-and-best-practices)".

To begin with, we need a data set, so we construct one.

```{r}
dataset <- data.frame(
  names = c("Betty Davis", "Peggy Sue", "Frank Zappa"),
  dob = c("1944-07-26", "1957-09-20", "1940-12-21"),
  gender = c("F", "F", "M"),
  v1 = sample.int(50, 3),
  v2 = sample.int(50, 3),
  v3 = sample.int(70, 3)
) 

```

Now we have a data set containing a number of demographics (names, date of birth, gender) and some research data (v1-v3).

```{r, comment=NA}
print(dataset)
```

We will now use 3 different methods to create pseudonyms for this data:

1.  counters
2.  random number generators
3.  hashing

Counter and RNG pseudonyms can be made before the actual data collection. Hashing pseudonyms can be made from the moment some data about the participants is available.

# Counter pseudonyms

For this method, we want to create pseudonyms of the form `<prefix><number>`. We do this by creating a series of numbers, and combine it with a predefined prefix (e.g. "PP"). If you make the pseudonyms before the data are collected, make sure to create enough.

```{r}
prefix <- "PP"                # The prefix we want to use 
n1 <- 1:999                   # Make a series of integers. 
n2 <- sprintf("%03d", n1)     # Add leading zero(s) to make them equal length
n3 <- paste0(prefix, n2)      # Add prefix
```

Once we have done that, it's just a matter of combining the data columns from the original data set with the newly created "id"-column containing the counter pseudonyms.

```{r}
counter <- dataset[,4:6]          # Make a dataframe containing the research data
counter$id <- n3[1:nrow(counter)] # Add the counter id column
dataset$id <- counter$id          # Add the counter id column to the keyfile
```

Presto! We have now a pseudonymized data set:

```{r, comment=NA}
print(counter)
```

And a keyfile:

```{r, comment=NA}
print(dataset)
```

# Random Number Generator (RNG) pseudonyms

Creating an RNG pseudonym is similar to creating a counter pseudonym, except that now the list with pseudonyms is randomized before they are assigned to the participants.

```{r}
n1 <- 1:999                       # Make a series of integers
n2 <- sprintf("%03d", n1)         # Beautify the numbers with leading zero
n3 <- paste("PP", n2, sep = "")   # Add prefix
n4 <- sample(n3)                  # Randomize the order of the pseudonyms
```

Again, we create a pseudonomyzed data set, as well as a keyfile

```{r}
rng <- dataset[,4:6]      # Make a dataframe containing the research data
rng$id <- n4[1:nrow(rng)] # Add the rng id column
dataset$id <- rng$id      # Add the rng id column to the keyfile
```

We have now a pseudonymized data set:

```{r, comment=NA}
print(rng)
```

And the keyfile:

```{r, comment=NA}
print(dataset)
```

# Hashed pseudonyms (with secret key)

In addition to counters and random number generators, hashing is also used in some cases to construct pseudonyms. Hashing is a process in which information (i.e. a combination of personal variables) is converted into a unique fixed-length code (the "hash" or "hash code") using a hashing algorithm.

What is unique about hashing is that it only works in one direction. You can transform information into a hash code via the hash function, but you cannot transform the hash code back to the original information.

Pseudonyms made by hashing are sometimes preferred because they allow you to generate the pseudonyms from the data, on the fly, meaning that in principle, no keyfile is necessary. You only need to know how the hashes are created and the key/secret that was used. For instance in longitudinal setups, this can be an advantage.

However, if not done properly, there is also a major threat: If no key is used while hashing it is possible to "guess" hash codes by hashing combinations of personal data and comparing the resulting codes. **Hashing without a key is therefore unsuitable for pseudonymization**. To securely generate hash codes, using a secret 'key' is indispensable.

## Step by step guide

In this example we use the `openssl` library for hashing, but the same results can be accomplished using the `digest` library.
```{r, message=FALSE}
library(openssl)
```

We start by choosing a key, which will be used as a kind of password while hashing. The key should be kept secret. It is therefore important to remove the key from your code from the moment that the pseudonyms are created. If you want to keep the key safe and failing memory proof, put it in a password manager for later use.
```{r}
MySecret = "this is the secret Key which you should REMOVE from you code"
```

The next step is choosing the personal data item you want to use to derive the hash codes from. Of course, for each person in the data set, this combination should be unique. Therefore, we chose to paste together the name, date of birth and gender into one long string.
```{r}
tmp <- paste0(dataset$names, dataset$dob, dataset$gender)   # Paste together personal data, create a unique string
```


Once we have done that, we pull these string through the hash-function (`sha2` from the `openssl`-library in this case) also using the secret key defined earlier.
```{r,echo=TRUE}
hashes <- sha2(tmp, key = MySecret)                        	# Hash it
```


In this step, we truncate (shorten) the resulting hash code in order to make them more manageable. This of course is an optional step. It is no problem at all to use the long hash codes. Even more, if you have **very big data sets** (say thousands of entries), **you should not truncate the hash codes** to avoid the risk of identical hash codes (i.e. collisions).
```{r}
hashes8 <- substr(hashes, 1, 8)      # Optionally truncate long hashes to something more workable. 
                                     # NOT for big data sets! 
```

Once the hashes are ready, we can create the pseudonymized data set containing the research data and the hash id's.
```{r}
hashed <- dataset[,4:6]
hashed$id <- hashes8
```

This is the resulting dataset:

```{r, comment=NA}
print(hashed)
```

Remember that it is not necessary to add the hash codes to the original data set. Instead, we have documented our method (with this code), and we can reveal the link between the data and the identities of the participants whenever it is necessary. In other words, it is no longer necessary to keep the research data available together with the personal data used to create the hashes.

Therefore, we only keep the data necessary to re-create the hash codes when necessary:

```{r, comment=NA}
keyfile <- subset(dataset, select = c(names, dob, gender))
```

and throw away the data set containing both the personal data and research data.

```{r}
rm(dataset)
```

Finally, we have two data sets left:

The pseudonymized data set containing the research data and the hash codes:

```{r, comment=NA}
print(hashed)
```

And the "keyfile" containing the variables used to create the hash codes:

```{r, comment=NA}
print(keyfile)
```

As the keyfile contains personal information, it should be kept in a safe place.
